\texttt{escheR}: Unified multi-dimensional visualizations with Gestalt principles

### 2023-10-05 Manuscript/manuscript.tex
Revision V2.0

\newcommand{\fixme}[1]{{\color{red} (#1)}}
\newcommand{\coloc}{\texttt{escheR}}

@@

%\setmainfont{Helvetica}
\title{\texttt{escheR}: Unified multi-dimensional visualizations with Gestalt principles}
\author[1]{Boyi Guo}
\author[2]{Louise A. Huuki-Myers}
\author[3,4]{Melissa Grant-Peters}
\author[2]{Leonardo Collado-Torres}
\author[1,5]{Stephanie C. Hicks\thanks{Correspondence to \url{shicks19@jhu.edu}}}
\affil[1]{Department of Biostatistics, Johns Hopkins Bloomberg School of Public Health, MD, USA}
\affil[2]{Lieber Institute for Brain Development, Johns Hopkins Medical Campus, Baltimore, MD, USA}
\affil[3]{Genetics and Genomic Medicine, Great Ormond Street Institute of Child Health, University College London, London, UK}
\affil[4]{Aligning Science Across Parkinson’s (ASAP) Collaborative Research Network, Chevy Chase, MD, USA}
\affil[5]{Malone Center for Engineering in Healthcare, Johns Hopkins University, MD, USA}
\date{\today}

@@

\subsection*{Availability and implementation}
The open source R package \texttt{escheR} is freely available on Bioconductor (\href{https://bioconductor.org/packages/escheR}{bioconductor.org/packages/escheR}).
% and Zenodo (\fixme{add link}).

@@

Here, we focus on two types of visualizations in biomedical research, but note that the proposed ideas could be extended beyond these applications: (i) embedding visualizations, which project data into some low-dimensional embedding or mathematical space (e.g. Principal Components Analysis (PCA) \cite{hotelling_1933}, $t$-distributed Stochastic Neighbor Embedding ($t$-SNE) \cite{vandermaaten_2008}, or Uniform Manifold Approximation and Projection (UMAP) \cite{becht_2019}) and (ii) \textit{in situ} visualizations \cite{dries_2021, Lewis_2021, odonoghue_2021}, which aim to visualize molecules captured from \textit{in situ} imaging or sequencing technologies where \textit{in situ} refers to `in its original place'. Both of these visualizations represent data in a 2D space and are motivated by recent advances in experimental technologies that profile molecules, including DNA, RNA, and proteins, at a single-cell or spatial resolution \cite{kashima_2020, moffitt_2022}. Some most popular technologies include single-cell/nucleus RNA-sequencing (sc/snRNA-seq) \cite{aldridge_2020} and \textit{in situ} spatially-resolved transcriptomics \cite{larsson_2021}.

A common and fundamental challenge with both of these visualizations is how to visualize multi-dimensional information in a 2D space. For example, in \textit{in situ} visualizations, we often want to create a spatial map to visualize a continuous (e.g. gene expression) or discrete (e.g. cell type or spatial domain) variable representing molecular information in the original spatial location. However, it is challenging to simultaneously visualize multi-dimensional data, such as information from disparate data domains (such as expression domain and spatial domain) or disparate data modalities (such as transcriptomics and proteomics) in the same plot. Currently, best practices for this include making two different plots displayed side-by-side (\textbf{Figure \textbf{\ref{fig:visual}A-B}}), one for gene expression and one for spatial domains. This creates cognitive gaps on how to associate the disparate information or how to interpret the biological findings of this multi-dimensional information regarding their (micro-)environment or colocalization. While interactive visualizations \cite{keller_2021, pardo_2022, sriworarat_2023} have the potential to mitigate this challenge, they are infeasible for scientific communications in static media, such as printed work. Developing a static and unified visualization that enables the simultaneous display of multiple dimensions of information is crucial for biomedical research.

\begin{figure*}[!t]
\begin{center}
\includegraphics[width=\textwidth]{Manuscript/figure/insitu.jpg}
\caption{\small \textbf{\texttt{escheR} enables multi-dimensional spatial visualizations following the Gestalt principles of design}. (\textbf{A-B}) The traditional visualization displays the colocalization plot of the expression of two genes \textit{EFNA5} and \texit{EPHA5} (\textbf{A}) and the spatial domains from the dorsolateral prefrontal cortex (DLPFC) in postmorteum human brain \cite{huukimyers_2023} (\textbf{B}) side-by-side, creating challenges to cognitively connecting colocalization status to spatial domains. (\textbf{C}) The watercolor effect enables displaying spatial domains by color-coding only outlines of circles. (\textbf{D}) \texttt{escheR} enables the multi-dimensional \textit{in situ} visualization that simultaneously displays the cortex layers and the colocalization status, substantially improving interpretability.}
\label{fig:visual} 
\end{center}
\end{figure*}

@@

To address these challenges, here we leverage the Gestalt (German for “unified whole”) principles for design \cite{todorovic_2008, palmer_1999} as a way to visualize multi-dimensional data in 2D visualizations. We focus on the two types of data visualizations previously introduced that are widely used in biomedical research: (i) embedding visualizations and (ii) \textit{in situ} visualizations. We provide an R package, \texttt{escheR}, implementing these ideas, which is built on the state-of-the-art data visualization framework \texttt{ggplot2} in the R programming language. Finally, we comment on how these ideas could be extended to other types of visualization in biomedical research.

@@

Here, we provide an open-source package called \texttt{escheR} (named after the graphic artist M.C. Escher) in the R programming language \cite{R}, leading to a simplified interface to navigate the implementation of the multi-dimensional visualization in 2D space. By adapting \texttt{ggplot2} standard, \texttt{escheR} can be seamlessly integrated into many popular spatial resolved toolboxes, such as \texttt{SpatialLIBD }\cite{pardo_2022}, \texttt{Seurat} \cite{hao_2021}, \texttt{Giotto} \cite{dries_2021} to name a few, and allow further theme customization with ease.

Next, we give two use cases to exemplify some utility of the proposed spatial visualization: (i) the spatially differential gene colocalization in the human dorsolateral prefrontal cortex (DLPFC) using spatial transcriptomics data \cite{huukimyers_2023}; (ii) multi-dimensional UMAP highlighting differential gene expression in data-driven cell clusters \cite{freytag_2020}.

@@

Adding a third dimension to 2D plots has been a long-standing challenge in visualization \cite{odonoghue_2010}. Our proposal addresses this fundamental challenge by introducing simple but effective design principles. Unlike computer-based interactive visualizations, the proposed visualization is free from any platform and technology restriction, creating an accessible and economic solution. In addition, the proposed visualization is easily scalable and hence can be applied to all types of spatially resolved data. Combining with a binning strategy, similar to \texttt{schex} \cite{freytag_2020}, to avoid possible overplotting, the proposed visualization can be also applied to visualized image-based spatially resolved data \cite{caicedo_2017}, in addition to aforementioned spot-based spatially resolved data (\textbf{Figure~\ref{fig:visual}}).

@@

@@

\begin{itemize}[nosep]
    \item \textbf{Boyi Guo}:  Conceptualization, Methodology, Software, Validation, Formal analysis, Investigation, Data Curation, Writing, Visualization
    \item \textbf{Louise A. Huuki-Myers}: Conceptualization, Software
    \item \textbf{Melissa Grant-Peters}: Conceptualization
    \item \textbf{Leonardo Collado-Torres}: Conceptualization, Software
    \item \textbf{Stephanie C. Hicks}: Conceptualization, Resources, Writing - Review \& Editing, Visualization, Supervision, Project administration, Funding acquisition
\end{itemize}

@@

\subsection*{Funding}
This project was supported by the National Institute of Mental Health R01MH126393 (BG, SCH), and CZF2019-002443 (SCH) from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community Foundation. All funding bodies had no role in the design of the study and collection, analysis, and interpretation of data and in writing the manuscript.

\subsection*{Acknowledgements}
We would like to acknowledge Nicholas J. Eagles, Kristen R. Maynard, Mina Ryten, Leon Di Stefano, and Lukas M. Weber (appearing in alphabetic order of last name) for their helpful comments, feedback and suggestions on \texttt{escheR} functionality. Nicholas J. Eagles and Kristen R. Maynard are employed by the Lieber Institute for Brain Development; Mina Ryten is employed by University College London; Leon Di Stefano and Lukas M. Weber are from the Johns Hopkins Bloomberg School of Public Health, Department of Biostatistics.

@@

\begin{itemize}[nosep]
    \item Boyi Guo (ORCiD: \href{https://orcid.org/0000-0003-2950-2349}{0000-0003-2950-2349})
    \item Louise A. Huuki-Myers (ORCiD: \href{https://orcid.org/0000-0001-5148-3602}{0000-0001-5148-3602})
    \item Melissa Grant-Peters (ORCiD: \href{https://orcid.org/0000-0003-0585-0971}{0000-0003-0585-0971})
    \item Leonardo Collado-Torres (ORCiD: \href{https://orcid.org/0000-0003-2140-308X}{0000-0003-2140-308X})
    \item Stephanie C. Hicks (ORCiD: \href{https://orcid.org/0000-0002-7858-0231}{0000-0002-7858-0231})
\end{itemize}

@@

% \clearpage 
\printbibliography
\bibliography{refs}

@@

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=\textwidth]{Manuscript/figure/embedding.jpg}
\caption{\small \textbf{\texttt{escheR} enables multi-dimensional embedding visualizations}. The gene expression of \textit{POMGNT1} among peripheral blood mononuclear cells \cite{PBMC} under the UMAP representation. (\textbf{A}) The \texttt{schex} R/Bioconductor package uses color-coded convex hulls to annotate data-driven cell types, creating confusion when interpreting hexagons in overlapping hulls. (\textbf{B}) \texttt{escheR} plots hexagon-specific membership to improve interpretability.}
\label{fig:embedding} 
\end{center}
\end{figure*}

>>>>

\newcommand{\rr}[1]{{\color{red} #1}}
\newcommand{\coloc}{\texttt{escheR}}

@@

\setlength{\bibsep}{0pt plus 0.3ex}

\renewcommand\Affilfont{\fontsize{9}{10.8}\itshape}   % change auth affil

%\setmainfont{Helvetica}
\title{\texttt{escheR}: Unified multi-dimensional visualizations with Gestalt principles}
\author[1]{Boyi Guo}
\author[2]{\rr{Louise A. Huuki-Myers}}
\author[3,4]{\rr{Melissa Grant-Peters}}
\author[2]{\rr{Leonardo Collado-Torres}}
\author[1,5,6,7]{Stephanie C. Hicks\thanks{Correspondence to \url{shicks19@jhu.edu}}}
\affil[1]{Department of Biostatistics, Johns Hopkins Bloomberg School of Public Health, MD, USA} 
\affil[2]{Lieber Institute for Brain Development, Johns Hopkins Medical Campus, Baltimore, MD, USA}
\affil[3]{Genetics and Genomic Medicine, Great Ormond Street Institute of Child Health, University College London, London, UK}
\affil[4]{Aligning Science Across Parkinson’s (ASAP) Collaborative Research Network, Chevy Chase, MD, USA}
\affil[5]{Department of Biomedical Engineering, Johns Hopkins School of Medicine, Baltimore, MD, USA}
\affil[6]{Center for Computational Biology, Johns Hopkins University, Baltimore, MD, USA}
\affil[7]{Malone Center for Engineering in Healthcare, Johns Hopkins University, MD, USA}

\date{\today}

@@

\subsection*{Availability and implementation}
The open source R package \texttt{escheR} is freely available on \rr{Bioconductor (\href{https://bioconductor.org/packages/escheR}{bioconductor.org/packages/escheR}).}
% and Zenodo (\fixme{add link}).

@@

Here, we focus on two types of visualizations in biomedical research, but note that the proposed ideas could be extended beyond these applications: (i) embedding visualizations, which project data into some low-dimensional embedding or mathematical space (e.g. Principal Components Analysis %(PCA)
\cite{hotelling_1933}, $t$-distributed Stochastic Neighbor Embedding ($t$-SNE) \cite{vandermaaten_2008}, or Uniform Manifold Approximation and Projection (UMAP) \cite{becht_2019}) and (ii) \textit{in situ} visualizations \cite{dries_2021, Lewis_2021, odonoghue_2021}, which aim to visualize molecules captured from \textit{in situ} imaging or sequencing technologies where \textit{in situ} refers to `in its original place'. Both of these visualizations represent data in a 2D space and are motivated by recent advances in experimental technologies that profile molecules, including DNA, RNA, and proteins, at a single-cell or spatial resolution \cite{kashima_2020, moffitt_2022}. Some most popular technologies include single-cell/nucleus RNA-sequencing (sc/snRNA-seq)
\cite{aldridge_2020} and \textit{in situ} spatially-resolved transcriptomics \cite{larsson_2021}. 
% 
% \vspace{-0.15in}

A common and fundamental challenge with both of these visualizations is how to visualize multi-dimensional information in a 2D space. For example, in \textit{in situ} visualizations, we often want to create a spatial map to visualize a continuous (e.g. gene expression) or discrete (e.g. cell type or spatial domain) variable representing molecular information in the original spatial location. However, it is challenging to simultaneously visualize multi-dimensional data, such as information from disparate data domains (such as expression domain and spatial domain) or disparate data modalities (such as transcriptomics and proteomics) in the same plot. Currently, best practices for this include making two different plots displayed side-by-side (\textbf{Figure \textbf{\ref{fig:visual}A-B}}), one for gene expression and one for spatial domains. This creates cognitive gaps on how to associate the disparate information or how to interpret the biological findings of this multi-dimensional information regarding their microenvironment or colocalization. While interactive visualizations \cite{keller_2021, pardo_2022, sriworarat_2023} have the potential to mitigate this challenge, they are infeasible for scientific communications in static media, such as printed work. Developing a static and unified visualization that enables the simultaneous display of multiple dimensions of information is crucial for biomedical research.

\begin{figure*}[!t]
\begin{center}
\includegraphics[width=\textwidth]{Manuscript/figure/insitu.jpg}
\caption{\small \textbf{\texttt{escheR} enables multi-dimensional spatial visualizations following the Gestalt principles of design}. (\textbf{A-B}) The traditional visualization displays the colocalization plot of the expression of two genes \textit{EFNA5} and \textit{EPHA5} (\textbf{A}) and the spatial domains from the dorsolateral prefrontal cortex in postmorteum human brain \cite{huukimyers_2023} (\textbf{B}) side-by-side, creating challenges to cognitively connecting colocalization status to spatial domains. (\textbf{C}) The watercolor effect enables displaying spatial domains by color-coding only outlines of circles. (\textbf{D}) \texttt{escheR} enables the multi-dimensional \textit{in situ} visualization that simultaneously displays the cortex layers and the colocalization status, substantially improving interpretability.}
\label{fig:visual} 
\end{center}
\end{figure*}

% \vspace{-0.2in}

@@

To address these challenges, here we leverage the Gestalt (German for “unified whole”) principles \cite{todorovic_2008, palmer_1999} as a path to visualize multi-dimensional data in 2D visualizations. We focus on the two types of data visualizations previously introduced that are widely used in biomedical research: (i) embedding visualizations and (ii) \textit{in situ} visualizations. We provide an R package, \texttt{escheR}, implementing these ideas, which is built on the state-of-the-art data visualization framework \texttt{ggplot2} in the R programming language. Finally, we comment on how these ideas could be extended to other types of visualization in biomedical research.

@@

Here, we provide an open-source package called \texttt{escheR} (named after the graphic artist M.C. Escher) in the R programming language \cite{R}, leading to a simplified interface to navigate the implementation of the multi-dimensional visualization in 2D space. By adapting \texttt{ggplot2} standard, \texttt{escheR} can be seamlessly integrated with \rr{popular data objects, including \texttt{SingleCellExperiment} \cite{amezquita2020} and \texttt{SpatialExperiment} \cite{righelli2022}, and easily work with popular pipelines, such as} \texttt{Seurat} \cite{hao_2021}, \texttt{Giotto} \cite{dries_2021} to name a few, and allow further theme customization with ease.

Next, we give two use cases to exemplify some utility of the proposed spatial visualization: (i) the spatially differential gene colocalization in the human dorsolateral prefrontal cortex using spatial transcriptomics data \cite{huukimyers_2023}; (ii) multi-dimensional UMAP highlighting differential gene expression in data-driven cell clusters \cite{freytag_2020}.

@@

Adding a third dimension to 2D plots has been a long-standing challenge in visualization \cite{odonoghue_2010}. Our proposal addresses this fundamental challenge by introducing simple but effective design principles. \rr{These principles lead to visually easier-to-interpret graphics compared to 2D-color gradients and geometry annotations (\textbf{Figure~\ref{fig:embedding}}).} Unlike computer-based interactive visualizations, the proposed visualization is free from any platform and technology restriction, creating an accessible and economical solution. In addition, the proposed visualization is easily scalable and hence can be applied to all types of spatially resolved data.

@@

\rr{
While the proposed design principle set-up a solid foundation, the proposed package \texttt{escheR} can be further optimized for image-based spatially resolved data \cite{caicedo_2017}. For example, to address possible overplotting of densely arranged single cells in tight space, strategies such as binning or hatching plots \cite{patrick2023spatial} can be incorporated. In addition, the flexibility to represent data points as polygons will be provided to allow visualization of cell morphology.
% Combining with a binning strategy, similar to \texttt{schex} \cite{freytag_2020} to avoid possible overplotting \rr{or hatching plot proposed in \texttt{lisaClust} \cite{patrick2023spatial}}, the proposed visualization can be also applied to visualized image-based spatially resolved data , in addition to aforementioned spot-based spatially resolved data (\textbf{Figure~\ref{fig:visual}}).  \rr{\texttt{escheR} could also be applied with kernel gene-weighted density plots from \texttt{Nebulosa} \cite{alquicirahernandez_2021} and other ggplot2-based visualizations.}
% \rr{Independently, \texttt{lisaClust} \cite{patrick2023spatial} visualize cell types within spatial domains by generating hatching plots for immunofluorescence and mass cytometry imaging data.} 
}

@@

\begin{itemize}[nosep]
    \item \textbf{Boyi Guo}:  Conceptualization, Methodology, Software, Validation, Formal analysis, Investigation, Data Curation, Writing, Visualization
    \item \textbf{Erik D. Nelson}: Conceptualization, Software, Writing, Visualization
    \item \textbf{Louise A. Huuki-Myers}: Conceptualization, Software
    \item \textbf{Melissa Grant-Peters}: Conceptualization
    \item \textbf{Leonardo Collado-Torres}: Conceptualization, Software
    \item \textbf{Stephanie C. Hicks}: Conceptualization, Resources, Writing - Review \& Editing, Visualization, Supervision, Project administration, Funding acquisition
\end{itemize}

@@

\subsection*{Funding}
This project was supported by the National Institute of Mental Health [R01MH126393 to BG and SCH, U01MH122849 to LAHM and LCT]; the Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community Foundation [CZF2019-002443 to SCH]; the Lieber Institute for Brain Development to LAHM and LCT; and Aligning Science Across Parkinson’s [ASAP-000478, ASAP-000509 to MGP] through the Michael J. Fox Foundation for Parkinson’s Research.  All funding bodies had no role in the design of the study and collection, analysis, and interpretation of data and in writing the manuscript.

\subsection*{Acknowledgments}
We would like to acknowledge Nicholas J. Eagles, Kristen R. Maynard, Mina Ryten, Leon Di Stefano, and Lukas M. Weber (appearing in alphabetic order of last name) for their helpful comments, feedback and suggestions on \texttt{escheR} functionality. Nicholas J. Eagles and Kristen R. Maynard are employed by the Lieber Institute for Brain Development; Mina Ryten is employed by University College London; Leon Di Stefano and Lukas M. Weber are from the Johns Hopkins Bloomberg School of Public Health, Department of Biostatistics.

@@

\begin{itemize}[nosep]
    \item Boyi Guo (ORCiD: \href{https://orcid.org/0000-0003-2950-2349}{0000-0003-2950-2349})
    \item Erik D. Nelson (ORCiD: \href{https://orcid.org/0000-0001-8477-0982}{0000-0001-8477-0982})
    \item Louise A. Huuki-Myers (ORCiD: \href{https://orcid.org/0000-0001-5148-3602}{0000-0001-5148-3602})
    \item Melissa Grant-Peters (ORCiD: \href{https://orcid.org/0000-0003-0585-0971}{0000-0003-0585-0971})
    \item Leonardo Collado-Torres (ORCiD: \href{https://orcid.org/0000-0003-2140-308X}{0000-0003-2140-308X})
    \item Stephanie C. Hicks (ORCiD: \href{https://orcid.org/0000-0002-7858-0231}{0000-0002-7858-0231})
\end{itemize}

@@

\clearpage 
% \printbibliography
\bibliography{refs}

@@

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=\textwidth]{Manuscript/figure/embedding.jpg}
\caption{\small \textbf{\texttt{escheR} \rr{improves} multi-dimensional embedding visualizations}. The gene expression of \textit{POMGNT1} among peripheral blood mononuclear cells \cite{PBMC} under the UMAP representation. (\textbf{A}) The \texttt{schex} R/Bioconductor package uses color-coded convex hulls to annotate data-driven cell types, creating confusion when interpreting hexagons in overlapping hulls. (\textbf{B}) \texttt{escheR} plots hexagon-specific membership to avoid substantial overlapping of cluster membership. \rr{Due to the binning strategy, cautions in interpretation are needed due to possible cluster intermixing within each hexagon.} }
\label{fig:embedding} 
\end{center}
\end{figure*}
